Measuring the double-drift illusion with hand trajectories

### 2022-01-05 Manuscript.Rmd
manuscript edits

We compare the illusory direction of the heading against the predictions of a simple vector combination model of perceived direction from Tse and Hsieh [-@Tse2006], Cavanagh and Tse [-@Cavanagh2019] and Heller et al. [-@Heller2021]. A more sophisticated model allows predicting perceived speed as well [@Johnston2013], but we are interested in illusion strength only, so we use the simple vector combination model. In this model, the perceived direction is a simple vector combination of the external ($V_e$) and internal ($V_i$) motions and a constant $K$. That is, the deviation from the physical direction is given by:

@@

The illusion cannot drift continuously away from the physical path forever. It can be disrupted by a temporal break [@Lisi2015] or by distracting attention [@Nakayama2020]. These disruptions may cause the perceived path to stay at a fixed offset, travelling parallel to the physical path, or it may return toward the veridical position before resuming the illusory direction (Fig. 3). Other percepts of resets have been reported as well, but as far as we know these fall in between these two extremes. Such resets may occur spontaneously once the accumulation has gone on too long or too far, depending on what mechanism causes the resets, and the strength of the illusion (purple arrows in Fig. 3). The illusion strength in turn in turn depends on the stimulus properties (speeds, gabor size, carrier frequency, eccentricity; @Gurnsey2012) as well as variations within and between participants. The purpose of this experiment is to capture this spontaneous resetting and determine its source.

@@

To facilitate modeling, the horizontal and vertical (straight ahead) coordinates of the reset points are kept isometric by giving them in centimeters. The distance from origin can then be converted to time as 1 _cm_ = 4/13.5 _s_ in the 4 _s_ condition and 1 _cm_ = 3/13.5 _s_ in the 3 _s_ condition.

@@

```{r fig5, echo=FALSE, message=FALSE, fig.width=6, fig.height=5.333333, fig.cap="\\label{fig:fig5}**Figure 5:** Reset models. **A:** Model limits relative to reset point locations. Reset points neither fall along a spatial limit (blue line; log likelihood: -1415.4) or a temporal limit (red lines: continuous for 4 _s_ passes, dashed for 3 _s_ passes; log likelihood: -837.0). Marginal distributions show that both the time and spatial offset of reset points approximate a gamma distribution. **B,C:** Color illustrates the gamma probability distributions, where white is zero probability, and the darkest red or blue is the maximum probability density. This is a relatively scale-free unit, where the total surface of the full probability density function is 1. The data and likelihood distributions match fairly well. **B:** Gamma distribution on time (log likelihood: -766.6). **C:** Gamma distribution on spatial offset (log likelihood: -1319.3)."}
invisible(capture.output(plotModels()))
```

>>>>

We compare the illusory direction of the heading against the predictions of a simple vector combination model of perceived direction from Tse and Hsieh [-@Tse2006], Cavanagh and Tse [-@Cavanagh2019] and Heller et al. [-@Heller2021]. A more sophisticated model allows predicting perceived speed as well [@Johnston2013], but this model predicts perceived direction and speed for a mixture of first order motions, and it is unclear if this works for a combination of first and second order motion (what we call internal and external speed here). In the vector average model we use here, the perceived direction is a simple vector combination of the external ($V_e$) and internal ($V_i$) motions and a constant $K$. That is, the deviation from the physical direction is given by:

@@

The illusion cannot drift continuously away from the physical path forever. It can be disrupted by a temporal break [@Lisi2015] or by distracting attention [@Nakayama2020]. These disruptions may cause the perceived path to stay at a fixed offset, travelling parallel to the physical path, or it may return toward the veridical position before resuming the illusory direction (Fig. 3). Other percepts of resets have been reported as well, but as far as we know these fall in between these two extremes. Such resets may occur spontaneously once the accumulation has gone on too long or too far, depending on what mechanism causes the resets, and the strength of the illusion (purple arrows in Fig. 3). The illusion strength in turn depends on the stimulus properties (speeds, gabor size, carrier frequency, eccentricity; @Gurnsey2012) as well as variations within and between participants. The purpose of this experiment is to capture this spontaneous resetting and determine its source.

@@

The X coordinates of the detected reset points are used as the spatial offset of the reset. We also need the time of the reset. Using the Y coordinates of the drawn trajectory (divided by speed) would assume that the real and perceived vertical position of the gabor are the same, which would imply that for very strong illusions the perceived speed can be many times the real speed. This seems unrealistic. As described above, it is unclear if the overall perceived speed of a mixture of many first order motions [@Johnston2013] applies to the double-drift illusion and if so, whether or not that can be used to calculate perceived time from path length. Instead, we chose a simplification by using the radius (R; distance from origin to reset, divided by speed) as the reset time. However, while useful for the current project, it remains to be tested how perceived speed and perceived path length are related to each other and to stimulus properties in the double-drift illusion.

To facilitate modeling, the horizontal and vertical (straight ahead) coordinates of the reset points are kept isometric by giving them in centimeters. The distance from origin can then be converted to time as 1 _cm_ = 4/13.5 _s_ in the 4 _s_ condition and 1 _cm_ = 3/13.5 _s_ in the 3 _s_ condition. For each model using a probability density function to estimate model likelihood, the variable that is used is divided by its median value such that each of their probability densities are on a comparable scale.

@@

```{r fig5, echo=FALSE, message=FALSE, fig.width=6, fig.height=5.333333, fig.cap="\\label{fig:fig5}**Figure 5:** Reset models. **A:** Model limits relative to reset point locations. Reset points neither fall close to a spatial limit (blue line) or a temporal limit (red line: 4 _s_ passes, orange line: 3 _s_ passes). Marginal distributions show that both the time and spatial offset of reset points approximate a gamma distribution. **B,C:** Color illustrates the gamma probability distributions, where white is zero probability, and the darkest red or blue is the maximum probability density. This is a relatively scale-free unit, where the total surface of the full probability density function (along a line orthogonal to the predictor variable) is 1. **B:** Gamma distribution on time (log likelihood: -605.6). **C:** Gamma distribution on spatial offset (log likelihood: -632.6)."}
invisible(capture.output(plotModels()))
```

When applied to the full dataset, the gamma distributions fit better than the normal distributions (normal space AIC=XXX, gamma space _AIC_=XXX; _p_<.XXX; normal time AIC=XXX, gamma time _AIC_=XXX; _p_<.XXX), but within neither of these there is a difference between the spatial and temporal distributions (normal _p_=.XXX; gamma _p_=.XXX). Therefore, we test how well each gamma distribution generalizes across sections of the data. For the spatial offset, we divide the data by Y coordinate in 5 equally sized bins. For the temporal offset, we divide the data by angle in 5 equally sized bins. We fit the gamma distribution to each bin. The fits within each bin are shown in Fig 5B&C. We then get likelihood estimates for the data in the 4 other bins, and sum these for a total estimate of how well a single distribution describes the whole data set. We find that the model on reset time generalizes better than the model on spatial reset offset (STATS).
